### 2014-10-09 PrivacyProtection-2014-09-09.tex
fixed some compilation errors and typos, the section EXESSIVE DATA MINING is now commented out

>\chapter{Legal and Ethical aspects of Privacy}
\label{chap:privacy}

@@

The term informational privacy as conceptual foundation of privacy originates from the juristic discussion The Right to Privacy (1890) by Warren and Brandeis.
The arise of new technologies like photography combined with a growing press created an increased publicity.
Warren and Brandeis were worried that existing law like castle doctrines or libel and slander could not suffice.
They felt a need for a new right to "protect the extent to wich one's thoughts, sentiments, and emotions could be shared with others" \cite{1}.
This right would be "the right to be left alone".

The right to "protect the extent to wich one's thoughts, sentiments, and emotions could be shared with others" by Warren and Brandeis was later extend to cover arbitrary information and is now kown as the privacy concept "... the control we have over information about ourselves ..." \cite{2}.

@@

``The European Court of Human has given this article a very broad
interpretation in its juresprudence.'' \cite{2}

\subsection{EU Data Protection Direvtive}

@@

EU directives in general do not hold any direct legal binding for EU citzens.
Member states have to create their own legal implementation.

@@

The sole responsibility for compliance is held by so called \textbf{controllers}.
A controller is an actor ``which alone or jointly with others dertimines the purposes and means of the processing of personal data; (art 2d)''
\cite{1}

@@

``Personal data should not be processed at all, except when certain conditions are met'' \cite{1}.
These conditions incorporate the seven OECD princepals and are futher categorized in the categories:

\begin{itemize}
\item
\textbf{Transparency}: ``The data subject has the right to be informed when his personal data is being processed [...] (art. 10 and 11)'' \cite{1} and he has the right how and by whom it is processed.
Additionally other speficied requirements have to be met, i.e.~giving consent.
\item
\textbf{Legitimate prupose}: ``Personal data can only be processed for specified explicit and legitimate pruposes and may not be processes futher in a way incompatible with those pruposes (art. 6b)''
\cite{1}.
\item
\textbf{Proportionality}: ``Personal data may be processed only insofar as it is adequate [...] in relation to [its] pruposes [...]; every reasonable step must be taken to ensure that data wich are inaccurate or incomplete [...] are erased or rectified (art. 6)'' \cite{1}
\end{itemize}

@@

Data transfer to thrid countries outside of the EU requires those countries to have a similar data protection level. Exceptions exists:

\begin{itemize}
\item
  Save Harbor
\item
  Passenger Name Record Aggreement
\end{itemize}

all between the US and the EU, although the US has no comparable law of data protection. 
The US Supreme Court regards privacy as \textbf{implicit} right granted with the First Amendment. However, european law states privacy as \textbf{explicit} right (ECHR Art. 8).
Additionally the US government follows a doctrine which favors the economy to implement self-regulation \cite{1}.

\subsection{Implementation of the Data Protection Direvtive}

@@

\item
The BDSG does not implement \textbf{independet} supervisory authorities. 
The \emph{Bundesdatenschutzbeauftragter} is subordinate to the Ministiry of Interior. 
Although he is not subject to technical oversight (\emph{Fachaufsicht}), he is subject to staff supervision by the government (\emph{Rechtsaufsicht} durch die Bundesregierung und  \emph{Dienstaufsicht} durch das Innenministerium) and budget oversight (including approval of employees) by the ministry \cite{7}. 
Thus the EC filed a treaty violation proceeding against Germany, again \cite{6}. 
In March 2010 Germany was found guilty of violation of Directive 95/46 by the ECJ \cite{6}.
\end{enumerate}

@@

The states of Germany have their own implementation of Directive 95/46 (\emph{Landesdatenschutzgesetze}).
Federal public authorites are only bound to their federal law. \cite{2}

@@

\begin{itemize}
\item
Right to disclosure of wether data about them is stored and processed, which data is stored and processed and the data sources.
\item
Right to correction of false data.

@@

The act is knwon for its high complexity: a manual record of phone numbers for business purposes could be hold subject to the DPA \cite{3}.
Although the act seems to fully cover the directive.
Even higher restriction apply for \emph{``sensitive personal data''} (race, ethnicity, politics, religion, trade union status, health, sex life or criminal record), i.e. \textbf{consent} must be given freely and has to be explicit. \cite{3}

@@

\item
View the data an organisation holds on them.
A `subject access request' can be obtained for a nominal fee. As of January 2014, the maximum fee is £2 for requests to credit reference agencies, £50 for health and educational request, and £10 per individual otherwise.

@@

France implements Directive 95/46 with Law 2004-801 modifying law 78-17 of 6.1.1978 (\emph{Loi n° 2004-801 du 6 août 2004} modifying \emph{loi n°78-17 relative à l'informatique, aux fichiers et aux libertés}) \cite{1}\cite{10}\cite{11}.

@@

\emph{Safe harbor} in general is the legal concept to regulate that a certain conduct will be deemed \cite{4}, but in germany the term is commonly used as synonym for the agreement between the EU and the US regarding Directive 95/46.)

@@

Many privacy researchers seem to ``forcus on the ways in which privacy can be infrenged'' \cite{1}.
Thus they try to create taxonomies of \emph{privacy harms} instead of taxonomies of \emph{privacy types}.
Those two differ in the respect that the former focuses on threats to prohibit whereas the latter focuses on values to protect.
So one should rather evaluate what aspects are precious about privacy and develop measures to ensure their security than only forbid single actions against privacy \cite{1}.

@@

Examples include:
\begin{itemize}
\item Body Characteristics: weight, height, sholder width, \ldots
\item Biometric Properties: Fingerprints, DNA Sequence, \ldots
\item Medical Conitions: Orthopedic conditions (e.g.~limping), Having a cold, \ldots
\end{itemize}

@@

Examples include:
\begin{itemize}
\item regular visit at church, backery, doctors
\item sexual habits
\item political activities
\end{itemize}

@@

This type is concerned with ``making sure that individuals's data is not automatically available to other individuals and organisations''. This is Informational Privacy in an intuitive sense.

@@

This type is the right ``to move about in public or semi-public space withoug being identified, tracked or monitored''.
Additionally this type is concerned with the protection of one's home and private places (``right to solitude''.

@@

This type is the right ``to associate with whomever [one] wishes, withoug being monitored''.
It is concerned with the protection against the automatic record of one's contacts.
It does not imply that one is monitored because of the associations.

Examples include:
\begin{itemize}
\item Friends
\item Joining of orgnaizations (e.g.~political parties)
\end{itemize}
\end{enumerate}

@@

\item
  \textbf{Privacy of The Person:} reflexive violation
\item
  \textbf{Privacy of Behaviour and Action:} none
\item
  \textbf{Privacy of Communication:} none
\item
  \textbf{Privacy of Data and Image:} none
\item
  \textbf{Privacy of Thoughts and Feelings:} Some psychological diseases
  (e.g.~depression) have physiological impact. Such physiological
  patterns could be detected.
\item
  \textbf{Privacy of Location and Space:} none
\item
  \textbf{Privacy of Association:} none
\end{itemiz*iPrivacy of Behaviour and Action}
he Privacy of Behaviour and Action is concerned with one' and social activities (religious, po...l, sexual, \ldots). Ifhis type is violated, following implicit violations are possible:
item
  \textbf{Privacy of The Person:} Religious practices emph include body
  modifications (e.g.~circumcision).
\item
  \textbf{Privacy of Behaviour and Action:} reflexive violation
\item
  \textbf{Privacy of Communication:} none
\item
  \textbf{Privacy of Data and Image:} none
\item
  \textbf{Privacy of Thoughts and Feelings:} Social activities in
  general depend on a certain intellectual attitude. Such an actvity is
  the expressions of such an attitude.
\item
  \textbf{Privacy of Location and Space:} none
\item
  \textbf{Privacy of Association:} Recording religious, politcal or
  sexual activities can reveal association with churches, political
  parties or sexual partners.
\end{itemize}

@@

The Privacy of Communication is concerned with not havin such
communication (corresponence or vis-a-vis) intercepted. This is very
broad type of privacy. Depending on the contens of the intercepted
communication every other type can be violated:

@@

\item
  \textbf{Privacy of The Person:} Communication about body
  characteristics.
\item
  \textbf{Privacy of Behaviour and Action:} Communicatio about social
  activities.
\item
  \textbf{Privacy of Communication:} reflexive violation
\item
  \textbf{Privacy of Data and Image:} Communication containing one's
  passwords or other sensitive data.
\item
  \textbf{Privacy of Thoughts and Feelings:} Communication of thoughts
  and feelings, e.g.~wiretapping a flirt or a catholic confession
  ritual.
\item
  \textbf{Privacy of Location and Space:} Interception of face-to-face
  communication is only possible if one's location and space is violated
  (wiretapping).
\item
  \textbf{Privacy of Association:} Communication about one's
  associations (family members, churches, etc.).
\end{itemize}

@@

The Privacy of Data and Image is concered with one's data not beeing
automatically availabel to others. This also is a very broad type of
privacy. Depending of the data or image contents erery other type can be
violated:

@@

\item
  \textbf{Privacy of The Person:} Frequently visited doctors can reveal
  certain medical conditions, if such doctors are known specialsts. In
  general it could imply ill-being.
\item
  \textbf{Privacy of Behaviour and Action:} Frequently visited places in
  general can reveal association and hence implies social activities.
\item
  \textbf{Privacy of Communication:} If one's location is knwon, it is
  possible to intercept (wiretap) one's communication. This also may
  violate the right to private spaces.
\item
  \textbf{Privacy of Data and Image:} If one's location is known, it is
  possible shoot pictures. This violates the right to one's image
  (\emph{``Recht am eigenen Bild''}).
\item
  \textbf{Privacy of Thoughts and Feelings:} Frequently visited persons
  may imply certain thoughts and feelings, e.g.~having a mistress.
\item
  \textbf{Privacy of Location and Space:} reflexive violation
\item
  \textbf{Privacy of Association:} Frequently visted places can reveal
  associations simply by searching in maps or yellow-pages.
\end{itemize}

@@

\item
  \textbf{Privacy of The Person:} Association with tatoo artists could
  imply having tatoos or other body modifications
\item
  \textbf{Privacy of Behaviour and Action:} Association with churches or
  poltical organizations could imply certain activities.
\item
  \textbf{Privacy of Communication:} none
\item
  \textbf{Privacy of Data and Image:} none
\item
  \textbf{Privacy of Thoughts and Feelings:} Association with churchs of
  political organizations could imply a certain intellectual attitude.
\item
  \textbf{Privacy of Location and Space:} none
\item
  \textbf{Privacy of Association:} reflexive violation
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The model is depictured in Figure \ref{fig:refmodel}.
It is organized in four views (round boexes) that contain a number of components (rectangular boxes).

@@

\item \textbf{IT-Systems.}
The releveant IT-Systems under study. This encompasses hardware (e.g. servers, network infrastructure), as well as software and third party services.
The level of garnularity has to be detailed enough to express all possible threats to the assets at stake.

@@

A typical conflict is the \emph{Criminal-User-Conflict}:
A user wants to keep control over their private data.
A criminal wants to gain money.
The possibility of selling private data (user profiles) to advertizers, renders both interests conflicting.

\item \textbf{Vulnerabilities.}
All indentifiable weaknesses in the IT-System.

In the example of the cirminal-user-conflict, the criminal has to exploit a vulnerability, e.g. a weak password, to gain access to the private data abou the user.

@@

In the case that, the loss can be quantified monetary, and the likelihood of occurence of a threat can
be modeled probabilistically, the risk is given by the procuct
\[ \text{risk} = \text{loss} \cdot P[\text{threat}]. \]
In practice such a quantitative risk evaluation is often not possible, and a qualitative, heursitic, analysis is performed instead.

\item \textbf{Security Requirements.}
A set of interactions (e.g. threats) that shall not occure within the system in order to acieve its business objectives.
Security requirements are targeted to protect one or more assets.

@@

The \emph{concept view} is a realization of the the potential view of the system.
It specifies important interactions that require further planning.
It contians the following components:
\begin{itemize}
\item \textbf{Business Model.}
The plan to achieve business objectives.

\item \textbf{Accident Scenario.}
A concrete outline of an interaction that leads to an accident.
In particular the asset under threatened and expolited vulnerability need to be described.

\item \textbf{Attack Technique.}
A specific technique or technology to attack IT-Systems (Man in the Middel, Phishing, etc.).
In particular the attacking actor, the conflict of interest and the expolited vulnerability need to be described.

\item \textbf{Security Measures.}
It describes a plan of sufficient measures to secure the intended interactions and to aviod the unintended interactions.
Each security mesaure targets a vulnearbility of the system in order to reduce a risk for a certain thread.
\end{itemize}

@@

At first, one has to outline the current state of the system under study. This includes description of:
\begin{itemize}
\item all \textbf{Assets} which must be protected
\item all relevant \textbf{IT-Systems}
\item all involved \textbf{Humans} and their \textbf{Conflicts of Interests}
\item all known \textbf{Vulnerabilities}
% \item and all important \textbf{Interactions} between the former components.
\end{itemize}

@@

Secondly, one needs to outline the potential insteractions of the system under study.
This includes both the unintended interactions (threads) and the intended interactions (business objectives).
This step produces four artefacts:
\begin{itemize}
\item a \emph{threat specifiation}, which identifies the threat, its targeted assets, the involved actors and thier conflicts of interest.
\item a \emph{threat risk evaluation}, which quantifies the likelyhood of a threat manifestation in relation to its associated loss.
\item a \emph{security requirement specifiation}, which specifies requirments in order to deal with identified hazards
\end{itemize}

@@

Based on \textbf{Step 2.}, the identified hazards are used alongside realistic accident ccenarios and attack techniques to create a \emph{risk matrix}.
With this matrix it is possible to decide if the risk is acceptable or not.
Together with the previously specified security requirements, the matrix is used to define adequate security measures.
Like a business model is an abstract concept to achieve business objectives, this step creates an concept to improve the system's security.

@@

The Reference Model can be used on different levels of abstraction.
This means each component can be used within a wide range of granularity, for instance the security measure \emph{Encryption} can be explored in genral or on the level of different concrete encryption tools; or on the even finer level of concrete algorithms.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pagebreak

@@

%TODO: UPDATE
Our definition of privacy is descibed in detail in Chapter \ref{chap:privacy}. In Section \ref{sec:taxonomy} we define privacy as the ``control over private data'' and introduce the following seven different types of privacy:
\begin{enumerate}
\item Privacy of the Person
\item Privacy of Behaviour and Action
\item Privacy of Communication
\item Privacy of Data and Image
\item Privacy of Thoughts and Feelings
\item Privacy of Location and Space
\item Privacy os Association
\end{enumerate}

@@

%\includegraphics{../diagrams/png/scenario-vulnerabilities.png}

%\paragraph{List of Vulnerabilities}
%
%The vulnerabilities of Live+Gov system outline are:
%
%\begin{itemize}
%\item
%  unencrypted data transmission (Man In The Middel)
%\item
%  insecure mobile devices (Exploit)
%\item
%  insecure servers (Exploit)
%\item
%  inadequate access rules for
%  \begin{itemize}
%  \item
%    mobile devices (citizens)
%  \item
%    servers (staff)
%  \item
%    applications (local authority)
%  \end{itemize}
%\item
%  corrupt/unhappy authorities or staff (\textbf{Could be actually a
%  threat not a vulnerability})
%\end{itemize}
%
%(\textbf{Note:} The vulnerabilities seem to mostly open possibilities to
%violate the \textbf{Privacy of Data and Image}, as this type is concerd
%with making data \emph{automatically} available to others.)

@@

\item
  \textbf{Unauthorized access to privacy sensitive data}, caused by
  \begin{itemize}
  \item
    \textbf{Excessive Data Mining} Linking sensor data provided by
    citizens with additional sources can produce more privacy sensitive
    data.
  \item
    \textbf{Corrupt Local Authorities} Local Authorities have certain data
    access, they could hand over this access for monetary reasons.
  \item
    \textbf{Corrupt/Unhappy Staff} Staff members also have a certain data
    access and they also could hand over this access for monetary reasons
    or as a form of payback for unfair treatment.
  \item
    \textbf{Competent Attackers} Competent Attackers are always a threat
    for IT-systems. Normaly they have monetary reasons to exploit a
    system, but there is also a possibility for proof-of-concept like
    attacks.
  \end{itemize}
\end{itemize}

\paragraph{Sensor Privacy Matrix (Figure \ref{figure:Live+Gov Sensor-Privacy Matrix})}

@@

The \textbf{Privacy of The Person} is generally concerned with one could
best understand as \emph{Biometric Privacy}. Friedewald et al. paraphase
it as \emph{``[...] the right to keep body functions and body
characteristics [...] private''}. \textbf{Accelerometer},
\textbf{Rotation Vector} and \textbf{Gyroscope} measure the physical
movement of the mobile device on all three axes. If the moblile device
is carried \emph{``normally''} its safe to say that those sensors also
measure the movments of its carrier. So his privacy is infringed
regarding biometric behaviour, as it is captured automatically.
(\textbf{Note:} This is not to be confused with the \textbf{Privacy of
Behaviour and Action} which is used for the social aspects of behaviour
action, e.g.~praying, sexual habits or poltical activities.)

@@

The \textbf{Privacy of Data and Image} demands, that
\emph{``individual's data is not automatically available to other
individuals''}. This type of privacy is trivially threatened because
here sensor data is indvidual data, a priori. So every sensor violates
the privacy of data and image, as data is transported into a foreign
system where operators have access to it.

@@

\emph{``Users should \textbf{have choices} regarding data collection
activies in terms of opting in or out, whether or not to provide data,
andhow to cerrect their data.''}

@@

\subparagraph{Consnet}

@@

This C might be special case of Choice. Before taking part a user should
have the choice wether to join or not (Opt-In).

>>>>

\chapter{Legal and Ethical aspects of Privacy}
\label{chap:privacy}

@@

The term informational privacy as conceptual foundation of privacy originates from the juristic discussion The Right to Privacy (1890) by Warren and Brandeis.
The arise of new technologies like photography combined with a growing press created an increased publicity.
Warren and Brandeis were worried that existing law like castle doctrines or libel and slander could not suffice.
They felt a need for a new right to "protect the extent to which one's thoughts, sentiments, and emotions could be shared with others" \cite{1}.
This right would be "the right to be left alone".

The right to "protect the extent to which one's thoughts, sentiments, and emotions could be shared with others" by Warren and Brandeis was later extend to cover arbitrary information and is now kown as the privacy concept "... the control we have over information about ourselves ..." \cite{2}.

@@

``The European Court of Human has given this article a very broad interpretation in its jurisprudence.'' \cite{2}

\subsection{EU Data Protection Directive}

@@

EU directives in general do not hold any direct legal binding for EU citizens.
Member states have to create their own legal implementation.

@@

The sole responsibility for compliance is held by so called \textbf{controllers}.
A controller is an actor ``which alone or jointly with others determines the purposes and means of the processing of personal data; (art 2d)''
\cite{1}

@@

``Personal data should not be processed at all, except when certain conditions are met'' \cite{1}.
These conditions incorporate the seven OECD principals and are further categorized in the categories:

\begin{itemize}
\item
\textbf{Transparency}: ``The data subject has the right to be informed when his personal data is being processed [...] (art. 10 and 11)'' \cite{1} and he has the right how and by whom it is processed.
Additionally other specified requirements have to be met, i.e.~giving consent.
\item
\textbf{Legitimate purpose}: ``Personal data can only be processed for specified explicit and legitimate purposes and may not be processes further in a way incompatible with those purposes (art. 6b)''
\cite{1}.
\item
\textbf{Proportionality}: ``Personal data may be processed only insofar as it is adequate [...] in relation to [its] purposes [...]; every reasonable step must be taken to ensure that data which are inaccurate or incomplete [...] are erased or rectified (art. 6)'' \cite{1}
\end{itemize}

@@

Data transfer to third countries outside of the EU requires those countries to have a similar data protection level. Exceptions exists:

\begin{itemize}
\item
  Save Harbor
\item
  Passenger Name Record Agreement
\end{itemize}

all between the US and the EU, although the US has no comparable law of data protection. 
The US Supreme Court regards privacy as \textbf{implicit} right granted with the First Amendment. However, European law states privacy as \textbf{explicit} right (ECHR Art. 8).
Additionally the US government follows a doctrine which favors the economy to implement self-regulation \cite{1}.

\subsection{Implementation of the Data Protection Directive}

@@

\item
The BDSG does not implement \textbf{independent} supervisory authorities. 
The \emph{Bundesdatenschutzbeauftragter} is subordinate to the Ministry of Interior. 
Although he is not subject to technical oversight (\emph{Fachaufsicht}), he is subject to staff supervision by the government (\emph{Rechtsaufsicht} durch die Bundesregierung und  \emph{Dienstaufsicht} durch das Innenministerium) and budget oversight (including approval of employees) by the ministry \cite{7}. 
Thus the EC filed a treaty violation proceeding against Germany, again \cite{6}. 
In March 2010 Germany was found guilty of violation of Directive 95/46 by the ECJ \cite{6}.
\end{enumerate}

@@

The states of Germany have their own implementation of Directive 95/46 (\emph{Landesdatenschutzgesetze}).
Federal public authorities are only bound to their federal law. \cite{2}

@@

\begin{itemize}
\item
Right to disclosure of whether data about them is stored and processed, which data is stored and processed and the data sources.
\item
Right to correction of false data.

@@

The act is known for its high complexity: a manual record of phone numbers for business purposes could be hold subject to the DPA \cite{3}.
Although the act seems to fully cover the directive.
Even higher restriction apply for \emph{``sensitive personal data''} (race, ethnicity, politics, religion, trade union status, health, sex life or criminal record), i.e. \textbf{consent} must be given freely and has to be explicit. \cite{3}

@@

\item
View the data an organization holds on them.
A `subject access request' can be obtained for a nominal fee. As of January 2014, the maximum fee is \pounds2 for requests to credit reference agencies, \pounds50 for health and educational request, and \pounds10 per individual otherwise.

@@

France implements Directive 95/46 with Law 2004-801 modifying law 78-17 of 6.1.1978 (\emph{Loi n2004-801 du 6 ao\^{u}t 2004} modifying \emph{loi n78-17 relative à l'informatique, aux fichiers et aux libert\'{e}s}) \cite{1}\cite{10}\cite{11}.

@@

\emph{Safe harbor} in general is the legal concept to regulate that a certain conduct will be deemed \cite{4}, but in Germany the term is commonly used as synonym for the agreement between the EU and the US regarding Directive 95/46.)

@@

Many privacy researchers seem to ``focus on the ways in which privacy can be infringed'' \cite{1}.
Thus they try to create taxonomies of \emph{privacy harms} instead of taxonomies of \emph{privacy types}.
Those two differ in the respect that the former focuses on threats to prohibit whereas the latter focuses on values to protect.
So one should rather evaluate what aspects are precious about privacy and develop measures to ensure their security than only forbid single actions against privacy \cite{1}.

@@

Examples include:
\begin{itemize}
\item Body Characteristics: weight, height, shoulder width, \ldots
\item Biometric Properties: Fingerprints, DNA Sequence, \ldots
\item Medical Conditions: Orthopedic conditions (e.g.~limping), Having a cold, \ldots
\end{itemize}

@@

Examples include:
\begin{itemize}
\item regular visit at church, bakery, doctors
\item sexual habits
\item political activities
\end{itemize}

@@

This type is concerned with ``making sure that individual's data is not automatically available to other individuals and organizations''. This is Informational Privacy in an intuitive sense.

@@

This type is the right ``to move about in public or semi-public space without being identified, tracked or monitored''.
Additionally this type is concerned with the protection of one's home and private places (``right to solitude''.

@@

This type is the right ``to associate with whomever [one] wishes, without being monitored''.
It is concerned with the protection against the automatic record of one's contacts.
It does not imply that one is monitored because of the associations.

Examples include:
\begin{itemize}
\item Friends
\item Joining of organizations (e.g.~political parties)
\end{itemize}
\end{enumerate}

@@

\item
  \textbf{Privacy of The Person:} reflexive violation
\item
  \textbf{Privacy of Behaviour and Action:} none
\item
  \textbf{Privacy of Communication:} none
\item
  \textbf{Privacy of Data and Image:} none
\item
  \textbf{Privacy of Thoughts and Feelings:} Some psychological diseases
  (e.g.~depression) have physiological impact. Such physiological
  patterns could be detected.
\item
  \textbf{Privacy of Location and Space:} none
\item
  \textbf{Privacy of Association:} none
\end{itemize}

\paragraph*{Privacy of Behaviour and Action}
The Privacy of Behaviour and Action is concerned with one' and social activities (religious, po...l, sexual, \ldots). Ifhis type is violated, following implicit violations are possible:
\begin{itemize}
\item
  \textbf{Privacy of The Person:} Religious practices which include body
  modifications (e.g.~circumcision).
\item
  \textbf{Privacy of Behaviour and Action:} reflexive violation
\item
  \textbf{Privacy of Communication:} none
\item
  \textbf{Privacy of Data and Image:} none
\item
  \textbf{Privacy of Thoughts and Feelings:} Social activities in
  general depend on a certain intellectual attitude. Such an activity is
  the expressions of such an attitude.
\item
  \textbf{Privacy of Location and Space:} none
\item
  \textbf{Privacy of Association:} Recording religious, political or
  sexual activities can reveal association with churches, political
  parties or sexual partners.
\end{itemize}

@@

The Privacy of Communication is concerned with not havin such
communication (correspondence or vis-a-vis) intercepted. This is very
broad type of privacy. Depending on the contents of the intercepted
communication every other type can be violated:

@@

\item
  \textbf{Privacy of The Person:} Communication about body
  characteristics.
\item
  \textbf{Privacy of Behaviour and Action:} Communication about social
  activities.
\item
  \textbf{Privacy of Communication:} reflexive violation
\item
  \textbf{Privacy of Data and Image:} Communication containing one's
  passwords or other sensitive data.
\item
  \textbf{Privacy of Thoughts and Feelings:} Communication of thoughts
  and feelings, e.g.~wiretapping a flirt or a catholic confession
  ritual.
\item
  \textbf{Privacy of Location and Space:} Interception of face-to-face
  communication is only possible if one's location and space is violated
  (wiretapping).
\item
  \textbf{Privacy of Association:} Communication about one's
  associations (family members, churches, etc.).
\end{itemize}

@@

The Privacy of Data and Image is concerned with one's data not being
automatically available to others. This also is a very broad type of
privacy. Depending of the data or image contents every other type can be
violated:

@@

\item
  \textbf{Privacy of The Person:} Frequently visited doctors can reveal
  certain medical conditions, if such doctors are known specialsts. In
  general it could imply ill-being.
\item
  \textbf{Privacy of Behaviour and Action:} Frequently visited places in
  general can reveal association and hence implies social activities.
\item
  \textbf{Privacy of Communication:} If one's location is knwon, it is
  possible to intercept (wiretap) one's communication. This also may
  violate the right to private spaces.
\item
  \textbf{Privacy of Data and Image:} If one's location is known, it is
  possible shoot pictures. This violates the right to one's image
  (\emph{``Recht am eigenen Bild''}).
\item
  \textbf{Privacy of Thoughts and Feelings:} Frequently visited persons
  may imply certain thoughts and feelings, e.g.~having a mistress.
\item
  \textbf{Privacy of Location and Space:} reflexive violation
\item
  \textbf{Privacy of Association:} Frequently visited places can reveal
  associations simply by searching in maps or yellow-pages.
\end{itemize}

@@

\item
  \textbf{Privacy of The Person:} Association with tattoo artists could
  imply having tattoos or other body modifications
\item
  \textbf{Privacy of Behaviour and Action:} Association with churches or
  political organizations could imply certain activities.
\item
  \textbf{Privacy of Communication:} none
\item
  \textbf{Privacy of Data and Image:} none
\item
  \textbf{Privacy of Thoughts and Feelings:} Association with churches of
  political organizations could imply a certain intellectual attitude.
\item
  \textbf{Privacy of Location and Space:} none
\item
  \textbf{Privacy of Association:} reflexive violation
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

The model is depictured in Figure \ref{fig:refmodel}.
It is organized in four views (round boxes) that contain a number of components (rectangular boxes).

@@

\item \textbf{IT-Systems.}
The relevant IT-Systems under study. This encompasses hardware (e.g. servers, network infrastructure), as well as software and third party services.
The level of granularity has to be detailed enough to express all possible threats to the assets at stake.

@@

A typical conflict is the \emph{Criminal-User-Conflict}:
A user wants to keep control over their private data.
A criminal wants to gain money.
The possibility of selling private data (user profiles) to advertisers, renders both interests conflicting.

\item \textbf{Vulnerabilities.}
All identifiable weaknesses in the IT-System.

In the example of the criminal-user-conflict, the criminal has to exploit a vulnerability, e.g. a weak password, to gain access to the private data abou the user.

@@

In the case that, the loss can be quantified monetary, and the likelihood of occurrence of a threat can
be modeled probabilistically, the risk is given by the product
\[ \text{risk} = \text{loss} \cdot P[\text{threat}]. \]
In practice such a quantitative risk evaluation is often not possible, and a qualitative, heuristic, analysis is performed instead.

\item \textbf{Security Requirements.}
A set of interactions (e.g. threats) that shall not occur within the system in order to achieve its business objectives.
Security requirements are targeted to protect one or more assets.

@@

The \emph{concept view} is a realization of the the potential view of the system.
It specifies important interactions that require further planning.
It contains the following components:
\begin{itemize}
\item \textbf{Business Model.}
The plan to achieve business objectives.

\item \textbf{Accident Scenario.}
A concrete outline of an interaction that leads to an accident.
In particular the asset under threatened and exploited vulnerability need to be described.

\item \textbf{Attack Technique.}
A specific technique or technology to attack IT-Systems (Man in the Middle, Phishing, etc.).
In particular the attacking actor, the conflict of interest and the exploited vulnerability need to be described.

\item \textbf{Security Measures.}
It describes a plan of sufficient measures to secure the intended interactions and to aviod the unintended interactions.
Each security measure targets a vulnerability of the system in order to reduce a risk for a certain thread.
\end{itemize}

@@

At first, one has to outline the current state of the system under study. This includes description of:
\begin{itemize}
\item all \textbf{Assets} which must be protected
\item all relevant \textbf{IT-Systems}
\item all involved \textbf{Humans} and their \textbf{Conflicts of Interests}
\item all known \textbf{Vulnerabilities}
\item and all important \textbf{Interactions} between the former components.
\end{itemize}

@@

Secondly, one needs to outline the potential interactions of the system under study.
This includes both the unintended interactions (threads) and the intended interactions (business objectives).
This step produces four artifacts:
\begin{itemize}
\item a \emph{threat specification}, which identifies the threat, its targeted assets, the involved actors and their conflicts of interest.
\item a \emph{threat risk evaluation}, which quantifies the likelihood of a threat manifestation in relation to its associated loss.
\item a \emph{security requirement specification}, which specifies requirements in order to deal with identified hazards
\end{itemize}

@@

Based on \textbf{Step 2.}, the identified hazards are used alongside realistic accident scenarios and attack techniques to create a \emph{risk matrix}.
With this matrix it is possible to decide if the risk is acceptable or not.
Together with the previously specified security requirements, the matrix is used to define adequate security measures.
Like a business model is an abstract concept to achieve business objectives, this step creates an concept to improve the system's security.

@@

The Reference Model can be used on different levels of abstraction.
This means each component can be used within a wide range of granularity, for instance the security measure \emph{Encryption} can be explored in general or on the level of different concrete encryption tools; or on the even finer level of concrete algorithms.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pagebreak

@@

%TODO: UPDATE
Our definition of privacy is described in detail in Chapter \ref{chap:privacy}. In Section \ref{sec:taxonomy} we define privacy as the ``control over private data'' and introduce the following seven different types of privacy:
\begin{enumerate}
\item Privacy of the Person
\item Privacy of Behaviour and Action
\item Privacy of Communication
\item Privacy of Data and Image
\item Privacy of Thoughts and Feelings
\item Privacy of Location and Space
\item Privacy os Association
\end{enumerate}

@@

@@

%\begin{itemize}
%
%\item
%  \textbf{Unauthorized access to privacy sensitive data}, caused by
%  \begin{itemize}
%  \item
%    \textbf{Excessive Data Mining} Linking sensor data provided by
%    citizens with additional sources can produce more privacy sensitive
%    data.
%  \item
%    \textbf{Corrupt Local Authorities} Local Authorities have certain data
%    access, they could hand over this access for monetary reasons.
%  \item
%    \textbf{Corrupt/Unhappy Staff} Staff members also have a certain data
%    access and they also could hand over this access for monetary reasons
%    or as a form of payback for unfair treatment.
%  \item
%    \textbf{Competent Attackers} Competent Attackers are always a threat
%    for IT-systems. Normaly they have monetary reasons to exploit a
%    system, but there is also a possibility for proof-of-concept like
%    attacks.
%  \end{itemize}
%\end{itemize}
%
%
\paragraph{Sensor Privacy Matrix (Figure \ref{figure:Live+Gov Sensor-Privacy Matrix})}

@@

The \textbf{Privacy of The Person} is generally concerned with one could
best understand as \emph{Biometric Privacy}. Friedewald et al. paraphrase
it as \emph{``[...] the right to keep body functions and body
characteristics [...] private''}. \textbf{Accelerometer},
\textbf{Rotation Vector} and \textbf{Gyroscope} measure the physical
movement of the mobile device on all three axes. If the mobile device
is carried \emph{``normally''} its safe to say that those sensors also
measure the moments of its carrier. So his privacy is infringed
regarding biometric behaviour, as it is captured automatically.
(\textbf{Note:} This is not to be confused with the \textbf{Privacy of
Behaviour and Action} which is used for the social aspects of behaviour
action, e.g.~praying, sexual habits or political activities.)

@@

The \textbf{Privacy of Data and Image} demands, that
\emph{``individual's data is not automatically available to other
individuals''}. This type of privacy is trivially threatened because
here sensor data is individual data, a priori. So every sensor violates
the privacy of data and image, as data is transported into a foreign
system where operators have access to it.

@@

\emph{``Users should \textbf{have choices} regarding data collection
activities in terms of opting in or out, whether or not to provide data,
and how to correct their data.''}

@@

\subparagraph{Consent}

@@

This C might be special case of Choice. Before taking part a user should
have the choice whether to join or not (Opt-In).
